Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-04-18 WSSSPE2_report/WSSSPE2_review.tex
Sections 1-4: fiddling.

The Second Workshop on Sustainable Software for Science: Practice and
Experiences
(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
held on Sunday, 16 November 2014 in conjunction with the International
Conference for High Performance Computing, Networking, Storage and Analysis
(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
model of a general initial workshop,
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
which co-occurred with SC13, and a focused workshop,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
which was organized in July 2014 jointly with the SciPy 
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
the development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational, and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged the broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future practice for software development and usage in science and
engineering. The workshop included multiple mechanisms for participation,
encouraged team building around solutions, and identified risky solutions with
potentially transformative outcomes. It strongly encouraged participation of
early-career scientists, postdoctoral researchers, and graduate students, with
funds provided to the conference organizers by the Moore Foundation and the
National Science Foundation, to support the travel of potential participants who
would not otherwise be able to attend, and young participants and those from
underrepresented groups, respectively.  These funds allowed \todo{number}
additional participants to attend, and each was offered the chance to present a
lightning talk.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynotes} \label{sec:keynotes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\note{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes}.
%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
\href{http://tinyurl.com/mnenzms}{Abstracts and slides}}

@@

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom. The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable 
world-class research. Chue Hong's keynote\footnote{Slides available at: http://dx.doi.org/10.6084/m9.figshare.1243288}
 started by making the point of that the use of 
-- and reliance on -- software is pervasive in all areas of world-leading research, 
showing examples from disciplines as diverse as the humanities and high-energy 
physics, and quoting Kersten Kleese Van Dam of the Pacific Northwestern 
National Laboratory via a petition campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
at change.org, ``\emph{Today there are very few science areas left 
which do not rely on IT and thus software for the majority of their research work. 
More importantly key scientific advances in experimental and observational science 
would have been impossible without better software.}'' He also cited Daniel Katz, 
Software Infrastructure for Sustained Innovation(SI2) Program Director of the 
National Science Foundation, ``\emph{Scientific discovery and innovation are 
advancing along fundamentally new pathways opened by development of 
increasingly sophisticated software. Software is an integral enabler of 
computation, experiment and theory, and directly responsible for increased 
scientific productivity and enhancement of researchers' capabilities.}''

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning Talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\note{Lead: Choi. Volunteers welcome.
\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}

@@

\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
through Sustainable Software Architectures.}
Venters proposed that technically sustainable software is a composite, first-class, 
non-functional requirement (NFR) that is at a minimum a measure of a systems 
maintainability and extensibility but may also include other NFR such as efficiency 
(energy and cost), interoperability, portability, reusability, scalability, and usability. To 
achieve technically sustainable software Venters suggested that software 
architectures are fundamental as they are the primary carrier of system qualities 
(NFR), i.e., pre-system understanding, and influence how developers are able to 
understand, analyze, extend, test and maintain a software system, i.e., post-
deployment system understanding. In addition, Venters highlighted that the 
sustainability of software architectures also needs to be addressed to endure 
different types of change and evolution in order to mitigate architectural drift and 
erosion, and architectural knowledge vaporization.

>>>>

The Second Workshop on Sustainable Software for Science: Practice and
Experiences
(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
held on  16 November 2014 in conjunction with the International
Conference for High Performance Computing, Networking, Storage and Analysis
(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
model of a general initial workshop,
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
which co-occurred with SC13, and a focused workshop,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
which was organized in July 2014 jointly with the SciPy 
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
the development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational, and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged the broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future practice for software development and usage in science and
engineering. The workshop included multiple mechanisms for participation,
encouraged team building around solutions, and identified risky solutions with
potentially transformative outcomes. It strongly encouraged participation of
early-career scientists, postdoctoral researchers, and graduate students, with
funds provided to the conference organizers by the Moore Foundation and the
National Science Foundation (NSF), to support the travel of potential participants who
would not otherwise be able to attend, and young participants and those from
underrepresented groups, respectively.  These funds allowed \todo{number}
additional participants to attend, and each was offered the chance to present a
lightning talk.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynotes} \label{sec:keynotes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes}.
%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
\href{http://tinyurl.com/mnenzms}{Abstracts and slides}}
\end{comment}

@@

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom. The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable 
world-class research. Chue Hong's keynote\footnote{Slides available at: http://dx.doi.org/10.6084/m9.figshare.1243288}
 started by making the point of that the use of 
-- and reliance on -- software is pervasive in all areas of world-leading research, 
showing examples from disciplines as diverse as humanities and high-energy 
physics, and quoting Kersten Kleese Van Dam of the Pacific Northwestern 
National Laboratory via a petition campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
at change.org, ``\emph{Today there are very few science areas left 
which do not rely on IT and thus software for the majority of their research work. 
More importantly key scientific advances in experimental and observational science 
would have been impossible without better software.}'' He also cited Daniel Katz, 
Software Infrastructure for Sustained Innovation (SI2) Program Director of the 
NSF, ``\emph{Scientific discovery and innovation are 
advancing along fundamentally new pathways opened by development of 
increasingly sophisticated software. Software is an integral enabler of 
computation, experiment and theory, and directly responsible for increased 
scientific productivity and enhancement of researchers' capabilities.}''

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning Talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{Lead: Choi. Volunteers welcome.
\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}
\end{comment}

@@

\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
through Sustainable Software Architectures.}
Venters proposed that sustainable software is a composite, first-class, 
non-functional requirement (NFR) that is at a minimum a measure of a system's 
maintainability and extensibility, but may also include other NFRs such as efficiency 
(e.g., energy, cost), interoperability, portability, reusability, scalability, and usability. To 
achieve technically sustainable software, Venters suggested that software 
architectures are fundamental as they are the primary carrier of system   
NFRs, i.e., pre-system understanding; and influence how developers are able to 
understand, analyze, extend, test, and maintain a software system, i.e., 
post-deployment system understanding. In addition, Venters highlighted that   
sustainability of software architectures needs to be addressed to endure 
different types of change and evolution in order to mitigate architectural drift, 
erosion, and knowledge vaporization.
